### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
finalizada revisao ateh sintese de texto

From the official 124 words, 
26 of them (20.97\%)
have only one syllable,
85 (68.55\%) have two syllables,
and 13 (10.48\%) have three syllables.
No official word has four or more syllables.
In all the 124 tokens, there are 235 syllables (68 different),
and Table~\ref{tab:freqSyl} exhibits the 10 most frequent
syllables in the first, last, middle and all positions.
Middle-word syllables only occur 13 times,
and are all different, with the exception of 'la'
which occurs twice.
A complete list of the syllables and their frequencies (in different
positions) is in the \tttt{hsyls} variable of the \tttt{makeStatistics.py}
script.
A list of all words, grouped by their size in syllables,
and ordered alphabetically and by the number of letters,
is in the \tttt{hlsyl\_\_} variable.

@@

Vowel and consonant frequencies,
and comparisons of vowels against consonants,
are as shown in Table~\ref{freqLet}
for start, end and internal position.
The limited number of consonants favors
the naturalness of the language prosody
as it resembles babbling.

@@

Given the rules in Section~\ref{phonology}
(of 14 letters, 5 vowels, (C)V(n) phonemes,
forbidden ji, wu, wo, ti, nn, nm),
96 words are possible with 1 syllable,
8256 with 2 syllables, 710016 with 3 syllables.
In the incident words of the official dictionary,
no middle syllable end with the nasal consonant 'n'.

Possible sentence structures are unlimited.
For a noun or verb phrase, one might quantify the possibilities
by considering all words but the particles that are not classified
also as something else
(i.e. li, e, la, pi, a, o, anu, en, seme, mu)
and the prepositions that are not classified also as something
else (i.e. kepeken, lon, tan).
This yields $120-13=107 words$.
As discussed in Section~\ref{syntax},
I advocate that, although such words are classified in the dictionary,
they might be used indistinctly as nouns, adjectives,
verbs and adverbs.
Thus, one has 107 possibilities of noun and verb phrases with
one word, $107^2=11,449$ possibilities with two words,
$107^3=1,225,043$ with three words and so on.
Be $n$, $v$, $o$, and $p$ the number of words in the subject (noun), predicate
(verb), object (noun) and prepositional (see Sections~\ref{sec:rec} and~\ref{sec:add}) phrases of a sentence,
and assume that the sentence has at most one prepositional phrase,
one might quantify the possibilities using the formula:
$\delta = 107^n\times 107^v\times 10^o \times 5\times 107^p$,
where $5$ stands for the possible prepositions.
To account for the particles, one possibility is
to assume for simplicity that one might use none or one particle
(not li, e, la, pi, resulting in $8$ particles and $9$ possibilities)
at each phrase, yielding:
$\delta = 107^n\times 107^v\times 107^o \times 5\times 107^p\times 9^4$.
For example, assume $n=v=o=p=1$, then
$\delta=107^1\times 107^1\times 107^1 \times 5\times 107^1\times 9^4=
 4.300066\times 10^+12$,
i.e. more then 4 trillion possible
sentences with single word phrases (i.e. no adjectives or adverbs),
while allowing only one particle per phrase
and only one predicate phrase
(e.g. 'ona li moku e soweli lon supa').
The general case is implemented as a function
in the script \tttt{makeStatistics.py}~\cite{tokipona}.

>>>>

From the official 124 words, 
26 of them (20.97\%)
have only one syllable,
85 (68.55\%) have two syllables,
and 13 (10.48\%) have three syllables.
No official word has four or more syllables.
In all the 124 tokens, there are 235 syllables (68 different),
and Table~\ref{tab:freqSyl} exhibits the 10 most frequent
syllables in the first, last, middle and all positions.
Middle-word syllables only occur 13 times,
and are all different, with the exception of 'la'
which occurs twice.
A complete list of the syllables and their frequencies (in different
positions) is in the \tttt{tf.hsyls} variable after executing the \tttt{basic/stardardAll.py}
script.
A list of all words, grouped by their size in syllables,
and ordered alphabetically and by the number of letters,
is in the \tttt{tf.hlsyl\_\_} variable.

@@

Vowel and consonant frequencies,
and comparisons of vowels against consonants,
are as shown in Table~\ref{freqLet}
for start, end and internal positions.
The limited number of consonants favors
the language prosody to appear natural
as it resembles babbling.

@@

Within the rules exposed in Section~\ref{phonology}
(of 14 letters, 5 vowels, (C)V(n) phonemes,
forbidden ji, wu, wo, ti, nn, nm syllables),
96 words are possible with 1 syllable,
8256 with 2 syllables, 710016 with 3 syllables.
In the incident words of the official dictionary,
no middle syllable end with the nasal consonant 'n'.

Possible sentence structures are unlimited.
For a noun or verb phrase, one might quantify the possibilities
by considering all words but the particles that are not classified
also as something else
(i.e. li, e, la, pi, a, o, anu, en, seme, mu)
and the prepositions that are not classified also as something
else (i.e. kepeken, lon, tan).
This yields $120-13=107$ words.
As discussed in Section~\ref{syntax},
the author advocates that, although words have specific classes in the dictionary,
they might be used indistinctly as nouns, adjectives,
verbs and adverbs.
Thus, one has 107 possibilities of noun and verb phrases with
one word, $107^2=11,449$ possibilities with two words,
$107^3=1,225,043$ with three words and so on.
Be $n$, $v$, $o$, and $p$ the number of words in the subject (noun), predicate
(verb), object (noun) and prepositional (see Sections~\ref{sec:rec} and~\ref{sec:add}) phrases of a sentence,
and assume that the sentence has at most one prepositional phrase,
one might quantify the possibilities using the formula:
$\delta = 107^n\times 107^v\times 10^o \times 5\times 107^p$,
where $5$ stands for the possible prepositions.
To account for the particles, one possibility is
to assume for simplicity the use of at most one particle
(not li, e, la, pi, resulting in $8$ particles and $9$ possibilities)
at each phrase, yielding:
$\delta = 107^n\times 107^v\times 107^o \times 5\times 107^p\times 9^4$.
For example, assume $n=v=o=p=1$, then
$\delta=107^1\times 107^1\times 107^1 \times 5\times 107^1\times 9^4=
 4.300066\times 10^+12$,
i.e. more then 4 trillion possible
sentences with single word phrases (i.e. no adjectives or adverbs),
while allowing only one particle per phrase
and only one predicate phrase
(e.g. 'ona li moku e soweli lon supa').~\cite{tokipona}.

### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/vowels.tex
terminada passagem de rotinas das tabelas do artigo para classes
finalizada revisao ateh sintese de texto

\begin{table*}[h!]
\scriptsize
\begin{center}
\caption{Frequency of letters in Toki Pona.
            freq, freq\_I, freq\_L and freq\_M are
            the frequencies of the letters in any, initial, last and middle
            positions.
            The columns 'v' and 'c' that follow them are frequencies
            considering only vowels and consonants.
            The most frequent vowel is 'a' in any position,
            although it is more salient among words starting with a vowel
            and among the last letter of the words.
            For starting, ending and middle positions, the second most frequent
            vowel varies.
            Among the consonants, 'n' is the most frequent because it is
            the only consonant allowed in the last position and because
            almost 20\% of the words end with 'n'.
            On the initial position, 's' is the most frequent consonant,
            while in middle position 'l' is the most frequent consonant.
            Many other conclusions can be drawn from this table and are
            useful e.g. for exploring sonorities in poems.}\label{freqLet}
\begin{tabular}{  l | c   c   c | c   c   c | c   c   c | c   c   c  }
letter & freq  & v  & c  & freq\_I  & v  & c  & freq\_L  & v  & c  & freq\_M  & v  & c \\\hline
a & 16.35  & 33.19  & -  & 8.06  & 40.00  & -  & 29.03  & 35.64  & -  & 14.22  & 29.46  & - \\
e & 8.60  & 17.45  & -  & 2.42  & 12.00  & -  & 11.29  & 13.86  & -  & 10.78  & 22.32  & - \\
i & 11.53  & 23.40  & -  & 3.23  & 16.00  & -  & 20.97  & 25.74  & -  & 10.78  & 22.32  & - \\
o & 7.55  & 15.32  & -  & 4.03  & 20.00  & -  & 14.52  & 17.82  & -  & 6.03  & 12.50  & - \\
u & 5.24  & 10.64  & -  & 2.42  & 12.00  & -  & 5.65  & 6.93  & -  & 6.47  & 13.39  & - \\\hline
j & 2.10  & -  & 4.13  & 3.23  & -  & 4.04  & 0.00  & -  & 0.00  & 2.59  & -  & 5.00 \\
k & 6.29  & -  & 12.40  & 11.29  & -  & 14.14  & 0.00  & -  & 0.00  & 6.90  & -  & 13.33 \\
l & 9.22  & -  & 18.18  & 12.10  & -  & 15.15  & 0.00  & -  & 0.00  & 12.50  & -  & 24.17 \\\hline
m & 4.61  & -  & 9.09  & 10.48  & -  & 13.13  & 0.00  & -  & 0.00  & 3.88  & -  & 7.50 \\
n & 10.48  & -  & 20.66  & 6.45  & -  & 8.08  & 18.55  & -  & 100.00  & 8.19  & -  & 15.83 \\
p & 5.66  & -  & 11.16  & 11.29  & -  & 14.14  & 0.00  & -  & 0.00  & 5.60  & -  & 10.83 \\\hline
s & 6.29  & -  & 12.40  & 13.71  & -  & 17.17  & 0.00  & -  & 0.00  & 5.60  & -  & 10.83 \\
t & 3.14  & -  & 6.20  & 6.45  & -  & 8.08  & 0.00  & -  & 0.00  & 3.02  & -  & 5.83 \\
w & 2.94  & -  & 5.79  & 4.84  & -  & 6.06  & 0.00  & -  & 0.00  & 3.45  & -  & 6.67 \\
\end{tabular}\end{center}
\end{table*}

>>>>

\begin{table*}[h!]
\scriptsize
\begin{center}
\caption{Frequency of letters in Toki Pona. Columns
                    freq, freq\_I, freq\_L and freq\_M are
                    the frequencies of the letters in any, initial, last and middle
                    positions.
                    The columns 'v' and 'c' that follow them are frequencies
                    considering only vowels and consonants.
                    The most frequent vowel is 'a' in any position,
                    although it is more salient among words starting with a vowel
                    and among the last letter of the words.
                    For starting, ending and middle positions, the second most frequent
                    vowel varies.
                    Among the consonants, 'n' is the most frequent because it is
                    the only consonant allowed in the last position and because
                    almost 20\% of the words end with 'n'.
                    On the initial position, 's' is the most frequent consonant,
                    while in middle position 'l' is the most frequent consonant.
                    Many other conclusions may be drawn from this table and are
                    useful e.g. for exploring sonorities in poems.}\label{freqLet}
\begin{tabular}{  l | c   c   c | c   c   c | c   c   c | c   c   c  }
letter & freq  & v  & c  & freq\_I  & v  & c  & freq\_L  & v  & c  & freq\_M  & v  & c \\\hline
a & 16.35  & 33.19  & -  & 8.06  & 40.00  & -  & 29.03  & 35.64  & -  & 14.22  & 29.46  & - \\
e & 8.60  & 17.45  & -  & 2.42  & 12.00  & -  & 11.29  & 13.86  & -  & 10.78  & 22.32  & - \\
i & 11.53  & 23.40  & -  & 3.23  & 16.00  & -  & 20.97  & 25.74  & -  & 10.78  & 22.32  & - \\
o & 7.55  & 15.32  & -  & 4.03  & 20.00  & -  & 14.52  & 17.82  & -  & 6.03  & 12.50  & - \\
u & 5.24  & 10.64  & -  & 2.42  & 12.00  & -  & 5.65  & 6.93  & -  & 6.47  & 13.39  & - \\\hline
j & 2.10  & -  & 4.13  & 3.23  & -  & 4.04  & 0.00  & -  & 0.00  & 2.59  & -  & 5.00 \\
k & 6.29  & -  & 12.40  & 11.29  & -  & 14.14  & 0.00  & -  & 0.00  & 6.90  & -  & 13.33 \\
l & 9.22  & -  & 18.18  & 12.10  & -  & 15.15  & 0.00  & -  & 0.00  & 12.50  & -  & 24.17 \\\hline
m & 4.61  & -  & 9.09  & 10.48  & -  & 13.13  & 0.00  & -  & 0.00  & 3.88  & -  & 7.50 \\
n & 10.48  & -  & 20.66  & 6.45  & -  & 8.08  & 18.55  & -  & 100.00  & 8.19  & -  & 15.83 \\
p & 5.66  & -  & 11.16  & 11.29  & -  & 14.14  & 0.00  & -  & 0.00  & 5.60  & -  & 10.83 \\\hline
s & 6.29  & -  & 12.40  & 13.71  & -  & 17.17  & 0.00  & -  & 0.00  & 5.60  & -  & 10.83 \\
t & 3.14  & -  & 6.20  & 6.45  & -  & 8.08  & 0.00  & -  & 0.00  & 3.02  & -  & 5.83 \\
w & 2.94  & -  & 5.79  & 4.84  & -  & 6.06  & 0.00  & -  & 0.00  & 3.45  & -  & 6.67 \\
\end{tabular}\end{center}
\end{table*}
